---
title: "Infections of the body and the mind"
output:
  distill::distill_article:
    self_contained: false
editor_options: 
  chunk_output_type: inline
citation_url: https://mondpanther.github.io/memonomicslab/hoaxism.html
---

**by [Julio Amador Diaz Lopez](https://www.imperial.ac.uk/people/j.amador)^[Imperial College Business School], [Ralf Martin](https://www.imperial.ac.uk/people/r.martin)^[Imperial College Business School  & Centre for Economic Perforamnce, LSE], Laure de Preux, Andre Veiga** 


<!--html_preserve-->
  <!-- Global site tag (gtag.js) - Google Analytics -->
  <script async src="https://www.googletagmanager.com/gtag/js?id=UA-3928947-4"></script>
  <script>
    window.dataLayer = window.dataLayer || [];
    function gtag(){dataLayer.push(arguments);}
    gtag('js', new Date());
  
    gtag('config', 'UA-3928947-4');
  </script>
<!--/html_preserve-->


Last update:  `r format(Sys.time(), '%B %d , %Y - %H:%M ')`

```{r Notes,eval=FALSE,include=FALSE}
#We are using this: https://www.earthdatascience.org/courses/earth-analytics/get-data-using-apis/use-twitter-api-r/

#This is useful too: https://rtweet.info/

```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rtweet)
library(tidytext)

# plotting and pipes - tidyverse!
library(ggplot2)
library(dplyr)
library(gdata)



```





```{r load ts data,include=FALSE,echo=FALSE}
#`r ntweets'


scomb=read.csv("https://www.dropbox.com/s/hgyl5ff6ql8000m/scomblong.csv?dl=1")


#adf=readRDS("../results/adf.rda")

#scomb=scomb%>%merge(adf,by="date")
#scomb=scomb%>%mutate(scaler=ifelse(scaler<1,1,scaler))
#scomb=scomb%>%mutate(stweets=tweets*scaler) %>%mutate(hoaxsh=igtweets/tweets)

ntweets= round(sum(scomb$tweets) /10^6,2)
nhoax=sum(scomb$igtweets)

# save to github
write.csv(scomb,"data/hoaxshare_over_time.csv")

```



The coronavirus pandemic has changed everything overnight. Unfortunately, as the genetic sequence of the virus started to make its deadly journey through bodies around the world, in parallel a memetic sequence emerged in the minds of some people: the idea that COVID19 pandemic is not real and a hoax. Indeed, the worry is that the two infections exist in a symbiotic relationship with one helping to advance the survival and spread of the other. Here we report on our ongoing efforts to map the spread of memetic infection using Twitter. Since March 23 we have been sampling tweets mentioning the terms “corona” and/or “covid”. We have also been sampling timelines of twitter users tweeting about these terms allowing us to go back further in time. Currently, we have collected   `r ntweets` million  tweets.


How bad is the hoax infection and is it getting better or worse? To identify tweeters believing in the hoax (or promoting the hoax idea) we look for tweets with one of the following hastags:

- "#hoax"
- "#coronahoax"
- "#covidhoax"


Using hashtags instead of string searches of the same terms provides a good distinction between tweets who display support for hoaxsim vs tweets criticising hoaxism. Note that this is likely a conservative way of counting hoaxist tweets and in reality a larger fraction of tweets are from people supporting hoaxist ideas.

Below is a time series plot of the share of hoaxist tweets over our sample period.^[ Data is
[here](https://mondpanther.github.io/memonomicslab/data/hoaxshare_over_time.csv)]
 we report separate series for the US and UK. Assigning location to tweets is notoriously difficult as most users have switch off detailed location tracking. In the figure below we base location on the analysis of a free text field where users can write something about their whereabouts. In many cases this refers to known areas although the detail varies (e.g. London, UK vs the Universe). Often it also involves phantasy locations (e.g. Walhalla). Hence, our "other" category might include tweeters from either the UK or Us who have chosen not to reveal their location.



```{r tsplot,echo=FALSE,message=FALSE}
scomb=scomb%>% group_by(country) %>% arrange(country,date) %>% mutate(cumtweets=cumsum(tweets),cumigtweets=cumsum(igtweets),cumhoaxsh=cumigtweets/cumtweets,hoaxsh=igtweets/tweets)


#names(scomb)


library(lubridate)

scomb=mutate(scomb,date=as_datetime(as.character(date)),country=as.character(country))


start.date = ymd_hms("2020-01-06 00:00:00")
end.date   = as_datetime(now()+days(7)) #ymd_hms("2020-04-02 01:00:00")

scomb=scomb %>% mutate(date=as_date(date))
  
breaksweeks = seq(date(start.date), date(end.date), "1 week")
scomb  = scomb %>% mutate(week=cut(date, breaks=breaksweeks)) 

scombw=scomb %>% filter(is.na(week)==FALSE) %>% group_by(country,week)%>% 
         summarise_at(vars(tweets,igtweets), sum, na.rm = TRUE)  %>% mutate(hoaxsh=igtweets/tweets, week=ymd(as.character(week))) %>% ungroup()
  



```

## Daily share of hoax tweets in total covid related tweets

```{r plot daily,echo=FALSE}

library(scales)
tsplot=ggplot(scomb, aes(x = date,y=hoaxsh*100,color=country  )  )+geom_line()  + theme_minimal() + xlab("Time") +
     ylab("Share of hoax tweets [%]")  + xlab("Days") +
     scale_x_date(breaks = date_breaks("1 months"),labels = date_format("%m/%y"));tsplot

```

## Weekly share of hoax tweets in total covid related tweets

```{r plot weekly,echo=FALSE}
tsplot=ggplot(scombw, aes(x = week,y=hoaxsh*100,color=country  )  )+geom_line()  + theme_minimal() + xlab("Time") +
     ylab("Share of hoax tweets [%]")+  xlab("Weeks")+
     scale_x_date(breaks = date_breaks("1 months"),labels = date_format("%m/%y"));tsplot

```


## Cumulative share of hoax tweets in total covid related tweets

```{r plot cumulative,echo=FALSE}
tsplot=ggplot(scomb, aes(x = date,y=cumhoaxsh*100,color=country  )  )+geom_line()  + theme_minimal() + xlab("Time") +
     ylab("Cumulative Share of hoax tweets [%]") + xlab("Days") +
     scale_x_date(breaks = date_breaks("1 months"),labels = date_format("%m/%y"));tsplot

```





